 

### 2020-12-10 vertical-thermal-review.Rmd
work on auto numbering of figures, appendices, plus some content on biophysical gradients

@@

Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and mortality.
Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves.
However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
Resulting $T_{leaf}$'s exceed air temperature ($T_{air}$) under conditions of high solar radiation, low wind, and low stomatal conductance.
Differences are most pronounced when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become dramatically elevated above both $T_{air}$ and understory $T_{leaf}$.
While leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, emerging tree-ring evidence suggests that understory trees experience relatively greater stress from high $T_{air}$.
Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
In the meantime, our findings imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.

@@

**However, we lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecosystems (Fig. 1).**

@@

i.	How does the biophysical environment vary with height in forests? ("What aspects of biophysical envt? Microclimate and vertical plant structure/phenology/etc (ok yes, see this in fig 1)"?--Elsa:
So change to micrometeorological environment? Not biological just physical in this question, right?

@@

The biophysical environment, here defined to include the physical structure of the forest canopy (from understory to the canopy top) and associated microclimates, is in large part determined by the structure of the forest itself. 
Canopy foliage, which varies across forest types and seasonally, strongly shapes the understory light, wind, etc.
Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. **2**).

@@

Light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
(Fig. **2x**), such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al. 2019].
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]. 
This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature [@fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].
-	Sunflecks: Leaky et al. 2003--from Marielle, "This is another, higher resolution axis of the variability of microenvironments that is probably quite important to address. (Also see the papers by Robin Chazdon on temporal variability of light environments). Maybe we will want to add these sentences (above) and something about sunflecks to a separate section that specifically addresses sources of variation in forest structure & microenvironments (e.g., vertical, horizontal, seasonal, diurnal, and with disturbance)?"

**Wind speeds are also higher at the top of the canopy (Fig. **2x**).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [e.g., REFS; @mcgregor_tree_nodate] and open savannas (Curtis et al. 2018).
This results in higher boundary layer conductance ($g_{a}$) for canopy leaves. Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata;($g_{a}$) also depends on leaf traits, morphology and windspeed. Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a] thus having important implications for leaf thermal sensitivity.

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. **2x**).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019]. 
Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018, Tymen et al. 2017]. 
"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy".
However, similar maximum temperatures have been observed during the dry season in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
Minimum $T_{air}$ is also buffered by forest canopies under some conditions. 
Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
*(Add something about risk of freeze.)* Additionally, canopies also buffer effects of snow through intercepting heavy snow accumulution on below-canopy levels [@nakamuraForestsTheirCanopies2017; @storckMeasurementSnowInterception2002]

**Humidity also varies across the forest vertical profile, being generally higher in the understory (Fig. **2x**).**
- DETAILS - REFS
- @mcgregor_tree_nodate

@@

![**Figure 2. Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is old version from Ian McGregor’s in press paper, showing NEON data from SCBI. See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2. ](NEON_height_profiles/NEON_height_profile_temp.png)

![This is a draft illustration of contents for the intendend figure. We'll pick out one variable-month combination per panel, and show it for all the forested NEON sites.](NEON_height_profiles/NEON_height_profile_temp2.png)

**The strength of this buffering varies across forests, being influenced by both forest characteristics and the biophysical environment.**
First, buffering increases with canopy cover.
* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015]
*Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
Taller trees don’t necessarily increase buffering [@zellweger_seasonal_2019]. 
Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
The strength of buffering also varies with respect to geographic and climatic factors.
* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
* [@davis_microclimatic_2019]
However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]

@@

Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b]. Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@niinemetsWorldwideAnalysisWithincanopy2015].
Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al. 2020], leaf thermoregulation and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)

@@

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ (Rollinson et al., in review).
Similarly, our previously unpublished analysis shows that in a broadleaf deciduous forest in Virginia, understory trees in a Mid-Atlantic forest responded more negatively to warmer $T_{air}$ than did canopy trees (Fig. **4**; Appendix **S3**). 
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. **3?**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. **3**).
It is also possible that competition is accelerated under warmer temperatures.

![**Figure 4. Temperature sensitivity of tree growth (derived from tree rings) for understory versus canopy trees.** THIS IS A ROUGH PLACEHOLDER-- KAT WILL PLAY WITH POTENTIAL VERSIONS. IF OUR IN-PREP TREE RING ANALYSIS METHODS PAPER IS ACCEPTED AHEAD OF THIS, WE CAN HIGHLIGHT RESULTS FROM THAT, WHICH ARE MORE RELIABLE. Analysis methods described in Appendix S3. OUTLIER HERE IS SHADE-TOLERANT BEECH. Data from Helcoski et al. 2019.](SCBI_tree_rings/SCBIfig.png)

@@

Appendix S1. Methods for NEON vertical profiles

Appendix S2. Methods for literature review

Appendix S3. Methods for SCBI tree-ring analysis

>>>>

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}

# This chunk is for auto numbering of tables/figures

#figures
fig_schematic = "1"
fig_NEON_vertical =  "2"
fig_NEON_differences  = "3"

# SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)

#Appendices
appendix_NEON_methods <- "S1"
appendix_tealeaves_methods <- "S2"
appendix_lit_review_methods <- "S3"
appendix_SCBI_treering <- "S4"

appendix_NEON_methods_title<- "Methods for NEON vertical profiles"
appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
appendix_lit_review_methods_title<- "Methods for literature review"
appendix_SCBI_treering_title <- "Methods for SCBI tree-ring analysis"

#SI table numbers
SItable_example = "S1"

#SI table captions
SItable_example_legend <- "Table S1. ...."

#SI figure numbers
SIfig_example = "S1"

#SI figure captions
SIfig_example_legend <- "Figure S1. ...."

```

@@

Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and mortality.
Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves, particularly in forests with dense canopies.
However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
Resulting $T_{leaf}$'s exceed air temperature ($T_{air}$) under conditions of high solar radiation, low wind, and low stomatal conductance.
Differences are most pronounced when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become dramatically elevated above both $T_{air}$ and understory $T_{leaf}$.
While leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, emerging tree-ring evidence suggests that understory trees experience relatively greater stress from high $T_{air}$.
Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
In the meantime, our findings imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.

@@

**However, we lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecosystems (Fig. ` r fig_schematic`).**

@@

i.	How does the biophysical environment vary with height in forests? ("What aspects of biophysical envt? Microclimate and vertical plant structure/phenology/etc (ok yes, see this in fig ` r fig_schematic`)"?--Elsa:
So change to micrometeorological environment? Not biological just physical in this question, right?

@@

The biophysical environment, here defined to include the physical structure of the forest canopy (from understory to the canopy top) and associated microclimates, is in large part determined by the structure of the forest itself. 
We supplement review of the existing literature with a new analysis of data on vegetation structure and vertical profiles in microclimate from the U.S. National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`; Schimel et al. 2007].

Canopy foliage, which varies across forest types and seasonally, strongly shapes the understory light, wind, etc.
Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. `r fig_NEON_vertical`).

@@

Light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
(Fig. `r fig_NEON_vertical`), such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al. 2019].
In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]. 
This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature [@fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].
-	Sunflecks: Leaky et al. 2003--from Marielle, "This is another, higher resolution axis of the variability of microenvironments that is probably quite important to address. (Also see the papers by Robin Chazdon on temporal variability of light environments). Maybe we will want to add these sentences (above) and something about sunflecks to a separate section that specifically addresses sources of variation in forest structure & microenvironments (e.g., vertical, horizontal, seasonal, diurnal, and with disturbance)?"

**Wind speeds are also higher at the top of the canopy (Fig. `r fig_NEON_vertical`).**
*(fill in some specifics from the NEON analysis)*
Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [e.g., REFS; @mcgregor_tree_nodate] and open savannas (Curtis et al. 2018).
This results in higher boundary layer conductance ($g_{a}$) for canopy leaves. Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata;($g_{a}$) also depends on leaf traits, morphology and windspeed. Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a] thus having important implications for leaf thermal sensitivity.

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019]. 
Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018, Tymen et al. 2017]. 
"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy".
However, similar maximum temperatures have been observed during the dry season in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
Minimum $T_{air}$ is also buffered by forest canopies under some conditions. 
Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
*(Add something about risk of freeze.)* Additionally, canopies also buffer effects of snow through intercepting heavy snow accumulution on below-canopy levels [@nakamuraForestsTheirCanopies2017; @storckMeasurementSnowInterception2002]

**Humidity also varies across the forest vertical profile, being generally higher in the understory (Fig. `r fig_NEON_vertical`).**
- DETAILS - REFS
- @mcgregor_tree_nodate

@@

![**Figure 2. Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is just one of several+ intended sites. See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2. ](NEON_height_profiles/figures/profile_SCBI.png)

**The strength of this buffering varies across forests, being influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
First, buffering increases with canopy cover.
* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015]
*Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
Taller trees don’t necessarily increase buffering [@zellweger_seasonal_2019]. 
Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
The strength of buffering also varies with respect to geographic and climatic factors.
* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
* [@davis_microclimatic_2019]
However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]

![**Figure 3. Role of canopy permeability in shaping temperature buffering, from NEON data.** Current placeholder figure is Nidhi's first rough analysis. See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2. ](NEON_height_profiles/figures/differences_across_forests_temp.png)

@@

Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b]. Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@niinemetsWorldwideAnalysisWithincanopy2015].
Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al. 2020], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)

@@

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ (Rollinson et al., in review).
Similarly, our previously unpublished analysis shows that in a broadleaf deciduous forest in Virginia, understory trees in a Mid-Atlantic forest responded more negatively to warmer $T_{air}$ than did canopy trees (Fig. **4**; Appendix `r appendix_SCBI_treering`). 
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. **3?**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. **3**).
It is also possible that competition is accelerated under warmer temperatures.

![**Figure 4. Temperature sensitivity of tree growth (derived from tree rings) for understory versus canopy trees.** THIS IS A ROUGH PLACEHOLDER-- KAT WILL PLAY WITH POTENTIAL VERSIONS. IF OUR IN-PREP TREE RING ANALYSIS METHODS PAPER IS ACCEPTED AHEAD OF THIS, WE CAN HIGHLIGHT RESULTS FROM THAT, WHICH ARE MORE RELIABLE. Analysis methods described in Appendix S4. OUTLIER HERE IS SHADE-TOLERANT BEECH. Data from Helcoski et al. 2019.](SCBI_tree_rings/SCBIfig.png)

@@

Appendix `r appendix_NEON_methods`. `r appendix_NEON_methods_title`

Appendix `r appendix_tealeaves_methods`. `r appendix_tealeaves_methods_title`

Appendix `r appendix_lit_review_methods`. `r appendix_lit_review_methods_title`

Appendix `r appendix_SCBI_treering`. `r appendix_SCBI_treering_title`
